Computing in the Field of Rationals

### 2012-06-25 rational_ext.tex
_

This view of symmetry between the value fragment of the language and
the continuation fragment was refined by Curien and Herbelin
\cite{Curien:2000} who introduce the sequent-style
$\mu\tilde{\mu}$-calculus, that exhibits symmetries between values and
continuations and between call-by-value and call-by-name. The calculus
includes the type $A-B$ which is the dual of implication, i.e., a
value of type $A-B$ is a context expecting a function of type $A
\rightarrow B$. Alternatively a value of type $A-B$ is also explained
as a \emph{pair} consisting of a value of type $A$ and a continuation
of type $B$. This is to be contrasted with our interpretation of a
value of that type as \emph{either} a value of $A$ or a demand for a
value of type $B$. The duality between call-by-name and call-by-value
was further investigated by Selinger using control
categories~\cite{Selinger:2001:CCD:966910.966911}. The
$\mu\tilde{\mu}$-calculus was further analyzed and extended by
Wadler~\cite{Wadler:2003,DBLP:conf/rta/Wadler05}. The extension gives
no interpretation to the subtraction connective and like the original
symmetric calculus of Filinski, introduces a duality that relates sums
to products and vice-versa.

Previous work on \emph{information effects} established the logically
reversible computational model {{Pi}} (and its extension {{Pi0}})
wherein computation is information preserving. The computational model
is derived from the isomorphisms of finite-types, every computation
admits an adjoint that expresses the logical ``inverse'' of a
computation and is complete for combinational circuits. The term model
of {{Pi}} is a so-called commutative bimonoidal category, a term used
for the categorification of a commutative semiring.

The language {{Pi}} however lacked a natural representation for
first-class functions due to first-order nature of all of its
values. In the light of the work of duality of computation it was felt
that ``right model'' for expressing functions in {{Pi}} would also
express symmetries similar to those in the calculi of Filinski, Curien
and Herbelin and Wadler. The technical goal of this paper is to
present the duality of computation in the setting of {{Pi}}.

The extended language, which we call {{PiEE}}, does not have one De
Morgan style dualizing action $(\_)^\bot$, but two distinct dualizing
actions -- an additive duality characterized by negative types $-(\_)$
and a multiplicative duality characterized by fractional types
$1/(\_)$. These negative and fractional dualities must not confused
with the duality in the additive and multiplicative fragment of linear
logic with the dualizing action $(\-)^\bot$. Each dualizing operation
preserves the underlying connective {{-(a+b)<->(-a)+(-b)}} and
{{1/(a*b)<->1/a*1/b}}.

The types of {{PiEE}} thus correspond to the field of rational numbers
(with the important restriction of fields that there is no division by
zero). A direct consequence of this is that algebraic manipulations of
rationals can be given an operational interpretation and \emph{have
  computational content}.

Models of linear logic rely on *-autonomous categories which change
the monoidal tensor when dualizing. Compact closed categories preserve
the monoidal tensor on dualizing.  The term model of {{PiEE}}
corresponds to the \emph{categorification of a field}. The category is
compact closed over the additive monoid. In the multiplicative monoid,
since there is no division by zero, respects compact closed structure
only when a dual object is defined.

>>>>

Symmetry between the value fragment of the language and the
continuation fragment was refined by Curien and Herbelin
\cite{Curien:2000} who introduced the sequent-style
$\mu\tilde{\mu}$-calculus, that exhibits symmetries between values and
continuations and between call-by-value and call-by-name. The calculus
includes the type $A-B$ which is the dual of implication, i.e., a
value of type $A-B$ is a context expecting a function of type $A
\rightarrow B$. Alternatively a value of type $A-B$ is also explained
as a \emph{pair} consisting of a value of type $A$ and a continuation
of type $B$. This is to be contrasted with our interpretation of a
value of that type as \emph{either} a value of $A$ or a demand for a
value of type $B$. The duality between call-by-name and call-by-value
was further investigated by Selinger using control
categories~\cite{Selinger:2001:CCD:966910.966911}. The
$\mu\tilde{\mu}$-calculus was further analyzed and extended by
Wadler~\cite{Wadler:2003,DBLP:conf/rta/Wadler05}. The extension gives
no interpretation to the subtraction connective and like the original
symmetric calculus of Filinski, introduces a duality that relates sums
to products and vice-versa.

Previous work on \emph{information effects} established the logically
reversible computational model {{Pi}} (and its extension {{Pi0}})
wherein computation is inherently information preserving. The
computational model is derived from the isomorphisms of finite-types,
every computation admits an adjoint that expresses the logical
``inverse'' of a computation and is complete for combinational
circuits. The term model of {{Pi}} is a so-called commutative
bimonoidal category, i.e. it corresponds to the categorification of a
commutative semiring with two commutative monoidal structures
{{(0,+)}} and {{(*,1)}} with {{*}} distributing over {{+}}.

The language {{Pi}} however lacks a natural representation for
first-class functions due to first-order nature of all of its
values. In the light of the work of duality of computation it was felt
that ``right model'' for expressing functions in {{Pi}} would also
express symmetries similar to those in the calculi of Filinski, Curien
and Herbelin and Wadler. The technical goal of this paper is to
present such a symmetric dual extension in the setting of {{Pi}}.

The extended language, which we call {{PiEE}}, is interesting because
it does not have one De Morgan style dualizing action $(\_)^\bot$, but
two distinct dualizing actions -- an additive duality characterized by
negative types $-(\_)$ and a multiplicative duality characterized by
fractional types $1/(\_)$. Each dualizing operation preserves the
underlying connective {{-(a+b)<->(-a)+(-b)}} and
{{1/(a*b)<->1/a*1/b}}. Negative and fractional types have a clear and
intuitive semantics in the setting of {{Pi}}.

These negative and fractional dualities must not confused with
$(\-)^\bot$ duality in the additive and multiplicative fragment of
linear logic.  Models of linear logic rely on *-autonomous categories
which change the monoidal tensor when dualizing. Compact closed
categories preserve the monoidal tensor on dualizing.  The term model
of {{PiEE}} corresponds to the \emph{categorification of a field}. The
category is compact closed over the additive monoid. In the
multiplicative monoid, since there is no division by zero, respects
compact closed structure only when a dual object is defined.

The types of {{PiEE}} correspond to the \emph{field of rational
  numbers} (with the important restriction of fields that there is no
division by zero). Consequently algebraic manipulations of rationals
can be given an operational interpretation and \emph{have
  computational content}.
